---
title: "BST210 Final Project Appendix"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

**Group Number: **7

**Group Name: **Regression Heroes

**Group Members:** Ryan Wang, Stella Nam, Hongkai Wang

```{r}
library(tidyverse)
library(stringr)
library(viridisLite) 
```

```{r, echo = F}
dat <- read_csv("data/cancer_reg.csv")
ed <- read_csv("data/Education.csv")
states <- read_csv("data/50_states.csv") %>% 
    add_row(State="District of Columbia", Abbr="DC", `State Capital`="Washington", Region="East")
```

```{r, echo=F}
# get county names
dat <- dat %>% 
    mutate(county_name = str_split(Geography, ",") %>% map_chr(., 1))%>%
    mutate(state = str_extract(Geography, "[^,]+$")) %>% # regex to select everything after ','
    mutate(state = str_trim(state)) %>%
    left_join(states, by=c("state" = "State")) %>%
    filter(county_name != "Valdez-Cordova Census Area")

Encoding(dat$county_name) <- "UTF-8"
dat$county_name <- iconv(dat$county_name, "UTF-8", "UTF-8", sub='')
dat <- dat %>%
    mutate(county_name = ifelse(county_name == "Doa Ana County", "Dona Ana County", county_name))

# get education stats from 2016-2020 https://www.ers.usda.gov/data-products/county-level-data-sets/download-data/
ed <- ed %>%
    filter(grepl("2016", Attribute)) %>%
    rename(county_name = `Area name`) %>%
    pivot_wider(names_from = Attribute, values_from = Value)  %>%
    mutate(county_name = case_when(county_name == "La Salle County" & State != "TX" ~ "LaSalle County",
                                   county_name == "La Salle Parish" ~ "LaSalle Parish",
                                   TRUE ~ county_name))

dat <- ed %>%
    right_join(dat, by=c("county_name", "State" = "Abbr")) %>%
    distinct()
```

```{r, echo=F}
dat = dat %>% filter(MedianAge <= 100)

hist(dat$MedianAge, breaks = 100)
```


```{r, echo=F}
ed_missing <- setdiff(unique(ed$county_name), unique(dat$county_name))
filter(ed, county_name %in% ed_missing)
dat_missing <- setdiff(unique(dat$county_name), unique(ed$county_name))
filter(dat, county_name %in% dat_missing)
```

# Linear model fitting:


```{r}
#need to rename some of the columns 
dat = dat %>% mutate(no_highschool = `Percent of adults with less than a high school diploma, 2016-20`, highschool = `Percent of adults with a high school diploma only, 2016-20`, some_college = `Percent of adults completing some college or associate's degree, 2016-20`, college = `Percent of adults with a bachelor's degree or higher 2016-20`)
```



```{r}
mod1 = lm(data = dat, TARGET_deathRate ~ medIncome + MedianAge + PctWhite)
summary(mod1)
```


```{r}
mod1.1 = lm(data = dat, TARGET_deathRate ~ medIncome + PctWhite)
summary(mod1.1)
anova(mod1.1, mod1)
```


```{r}
mod1.2 = lm(data = dat, TARGET_deathRate ~ medIncome + PctWhite + MedianAge + medIncome*MedianAge)
summary(mod1.2)
anova(mod1.2, mod1)

mod1.3 = lm(data = dat, TARGET_deathRate ~ medIncome + PctWhite + MedianAge + PctWhite*MedianAge)
summary(mod1.3)
anova(mod1.3, mod1)
```


```{r}
mod1.4 = lm(data = dat, TARGET_deathRate ~ medIncome + PctWhite + PctBlack + PctAsian + PctOtherRace)
summary(mod1.4)
```


```{r}
cor(dat$PctAsian, dat$PctWhite)
cor(dat$PctBlack, dat$PctWhite)
cor(dat$PctOtherRace, dat$PctWhite)
cor(dat$PctBlack, dat$medIncome)
cor(dat$PctWhite, dat$medIncome)
```


```{r}
mod1.5 = lm(data = dat, TARGET_deathRate ~ medIncome + PctWhite + PctAsian + PctOtherRace)
summary(mod1.5)
anova(mod1.5, mod1.1)
```


```{r}
mod1.6 = lm(data = dat, TARGET_deathRate ~ medIncome)
summary(mod1.6)
mod1.6.1 = lm(data = dat, TARGET_deathRate ~ medIncome + I(medIncome ^2))
summary(mod1.6.1)
anova(mod1.6, mod1.6.1)

predict = data.frame(TARGET_deathRate = predict(mod1.6.1, dat), medIncome = dat$medIncome)

dat %>% ggplot(aes(medIncome, TARGET_deathRate)) + geom_point() + geom_smooth(method = "lm") + geom_line(data = predict, aes(medIncome,TARGET_deathRate ), color = "red")
```


### check if population is a good predictor. 
```{r}

mod1.7 = lm(data = dat, TARGET_deathRate ~ popEst2015)
summary(mod1.7)
mod1.7.1 = lm(data = dat, TARGET_deathRate ~ popEst2015 + I(popEst2015 ^2))
summary(mod1.7.1)
anova(mod1.7, mod1.7.1)

predict = data.frame(TARGET_deathRate = predict(mod1.7.1, dat), popEst2015 = dat$popEst2015)

dat %>% ggplot(aes(popEst2015, TARGET_deathRate)) + geom_point() + geom_smooth(method = "lm") + geom_line(data = predict, aes(popEst2015,TARGET_deathRate ), color = "red")
# heavily influenced by outlier. Not a good approach./
```




### check if types of health insurance is a good predictor. 
```{r}

mod1.8 = lm(data = dat, TARGET_deathRate ~ PctPrivateCoverageAlone + PctPublicCoverageAlone)
summary(mod1.8)

mod1.8.1 = lm(data = dat, TARGET_deathRate ~ PctPublicCoverageAlone + I(PctPublicCoverageAlone*PctPrivateCoverageAlone))
summary(mod1.8.1)
anova(mod1.8, mod1.8.1)

predict = data.frame(TARGET_deathRate = predict(mod1.8.1, dat), PctPublicCoverageAlone = dat$PctPublicCoverageAlone)

dat %>% ggplot(aes(PctPublicCoverageAlone, TARGET_deathRate)) + geom_point() + geom_smooth(method = "lm") + geom_line(data = predict, aes(PctPublicCoverageAlone,TARGET_deathRate ), color = "red")
# heavily influenced by outlier. Not a good approach./
```

## check for possibility of GAM models. 

```{r}

dat %>% ggplot(aes(county_name, TARGET_deathRate)) + geom_point() + geom_smooth(method = "loess")
dat %>% ggplot(aes(avgAnnCount, TARGET_deathRate)) + geom_point() + geom_smooth(method = "loess")
dat %>% ggplot(aes(PctSomeCol18_24, TARGET_deathRate)) + geom_point() + geom_smooth(method = "loess")
dat %>% ggplot(aes(PctMarriedHouseholds, TARGET_deathRate)) + geom_point() + geom_smooth(method = "loess")  
dat %>% ggplot(aes(PctEmpPrivCoverage, TARGET_deathRate)) + geom_point() + geom_smooth(method = "loess")
dat %>% ggplot(aes(BirthRate, TARGET_deathRate)) + geom_point() + geom_smooth(method = "loess")
dat %>% ggplot(aes(county_name, TARGET_deathRate)) + geom_point() + geom_smooth(method = "loess")

```


We now have this following core model: 

```{r}
mod1_core = lm(data = dat, TARGET_deathRate ~ medIncome + I(medIncome ^2)+ MedianAge*medIncome+ PctWhite + PctAsian + PctOtherRace)
summary(mod1_core)

```
#### adding education status onto the existing model. 

we have the percentage of people who doesn't have high school diploma, finished high school education, some college experience, and finished bachelor diploma. we can start to test out all four covariates in the model. 


```{r}
mod1_core_edu_1 = lm(data = dat, TARGET_deathRate ~ medIncome + I(medIncome ^2)+ MedianAge*medIncome + PctWhite + PctAsian + PctOtherRace + no_highschool + highschool + some_college + college)

summary(mod1_core_edu_1)
#observe hint for colinearity.
```

```{r}
cov(dat$highschool, dat$no_highschool)
cov(dat$highschool, dat$college)
cov(dat$highschool, dat$some_college)

cov(dat$college, dat$some_college)
```

we can see that there is high negative correlation between the percentage of people having a high school diploma, and percentage of people who finished college or above. Thus, it makes sense that the college covariate is not significant in itself. 


```{r}
mod1_core_edu_2 = lm(data = dat, TARGET_deathRate ~ medIncome + I(medIncome ^2)+ PctWhite + PctAsian + PctOtherRace + no_highschool + highschool)

summary(mod1_core_edu_2)
```

Now we observe that percentage of Asian has lost its significance. It is possible that the addition of education level added confounding to the existing model. let's check for correlation between the covariates again. 

```{r}
cov(dat$PctAsian, dat$highschool)
cov(dat$PctAsian, dat$college)
cov(dat$PctAsian, dat$some_college)
cov(dat$PctAsian, dat$no_highschool)
```

```{r}
modelcheck = lm(data = dat, TARGET_deathRate ~ PctPublicCoverageAlone + PctPrivateCoverageAlone+ PctPublicCoverageAlone*PctPrivateCoverageAlone)

summary(modelcheck)
```


## final linear model form
```{r}
lin_final = lm(data = dat, TARGET_deathRate ~ medIncome + I(medIncome ^2) + MedianAge + PctWhite  + PctOtherRace + no_highschool + highschool  + PctPublicCoverageAlone + PctPrivateCoverageAlone+ PctPublicCoverageAlone*PctPrivateCoverageAlone + PercentMarried + PercentMarried*MedianAge)

summary(lin_final)

write.csv(lin_final$model,"output.csv")
```
### simple linear model evaluation. 
Let's evaluate the residual diagnostic to confirm the model's validity:


```{r}
standardized_res = rstandard(lin_final)
scatter.smooth(standardized_res, main = "Linear Regression Model Standardized Residuals", ylab = "Standarized Residual Value")
qplot(standardized_res, binwidth = 0.2, main = "Distribution of Standarized Residuals", ylab = "Count", xlab = "Residual Value")

qqnorm(standardized_res, pch = 1, frame = FALSE)
qqline(standardized_res, col = "steelblue", lwd = 2)
```


We can also try some other linear model fitting methods. We haven't done anything with the GAM models yet. 

### b. Logistic/multinomial/ordinal regression

First, we can split the lung-cancer death rate into several categories to broadly access the healthcare system at each county. For example, we can artificially create three different categories in the death rate variable. Let's take a look at the death rate distribution. 
```{r}
#Create a sequence of 100 equally spaced numbers between -4 and 4
x <- seq(-4, 4, length=100)

#create a vector of values that shows the height of the probability distribution
#for each value in x
y <- dnorm(x)

#plot x and y as a scatterplot with connected lines (type = "l") and add
#an x-axis with custom labels
plot(x,y, type = "l", lwd = 2, axes = FALSE, xlab = "", ylab = "")
axis(1, at = -3:3, labels = c("-3s", "-2s", "-1s", "mean", "1s", "2s", "3s"))
```


```{r}
hist(dat$TARGET_deathRate, breaks = 100, main = "Distribution of Lung Cancer-Related Death Rate", xlab = "County Death Rate")


```

We can make three bins, deathrate < 150, 150 <= deathrate < 200, 200 <= deathrate, and use them as a proxy to the quality of lung cancer prevention and quality of cancer care for each US county. The normal distribution of lung cancer death rate also suggests that dividing outcomes into bins doesn't really benefit us with our primary goal. 

But let's do it anyway to test it out anyway:


```{r}
#create the three bins
dat = dat %>% mutate(multi = case_when(TARGET_deathRate < 150 ~ 1, TARGET_deathRate < 200 ~ 2, T ~ 3))
# 3 is bad quality lung cancer prevention, 2 is medium, 1 is good quality.  
```



```{r}
library(nnet)
mod2.1 <- multinom(multi ~ medIncome + I(medIncome ^2), data = dat)
summary(mod2.1)
```

```{r}
plot(mod2.1$fitted.values[,1][order(dat$medIncome)] ~ sort(dat$medIncome), type="l", col="dodgerblue", xlab=c("Median Income"), ylab="Predicted Probability", ylim=c(0,1))
points(mod2.1$fitted.values[,2][order(dat$medIncome)] ~ sort(dat$medIncome), type="l", col="magenta")
points(mod2.1$fitted.values[,3][order(dat$medIncome)]~sort(dat$medIncome), type="l", col="green")
```



```{r}
mod2.2 <- multinom(multi ~ PctWhite , data = dat)
summary(mod2.2)


plot(mod2.2$fitted.values[,1][order(dat$PctWhite)] ~ sort(dat$PctWhite), type="l", col="dodgerblue", xlab=c("Percentage of white residents"), ylab="Predicted Probability", ylim=c(0,1))
points(mod2.2$fitted.values[,2][order(dat$PctWhite)] ~ sort(dat$PctWhite), type="l", col="magenta")
points(mod2.2$fitted.values[,3][order(dat$PctWhite)]~sort(dat$PctWhite), type="l", col="green")
```

### multinomial distribution of medical resources by states

```{r}
dat = dat%>%mutate(Region = case_when(state == "District of Columbia" ~ "Northeast", T~ Region))

```


```{r}
library(nnet)

mod3.0 = multinom(data = dat, Region ~ medIncome + MedianAge + popEst2015 + PctWhite + PctOtherRace + PctPublicCoverage + PctPrivateCoverage)

#South used as the baseline here. 
summary(mod3.0)

coef(mod3.0)
write.csv(coef(mod3.0), file = "multi.csv")
```


```{r}
regions = dat %>% select(state, Region) %>% unique()
write.csv(regions, file = "regions.csv")
```

```{r}
upper = coef(mod3.0) + 1.96*summary(mod3.0)$standard.errors
lower = coef(mod3.0) - 1.96*summary(mod3.0)$standard.errors

upper
lower 
coeff = coef(mod3.0)

```



```{r}

tmp = as.data.frame(cbind(mod3.0$fitted.values[,1], dat$MedianAge))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,2], tmp))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,3], tmp))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,4], tmp))

names = c("Northeast",        "South"        ,"West" , "Midwest","age")
tmp = setNames(tmp, names)
tmp = tmp %>% pivot_longer(cols = !age, names_to = c("region" ), values_to = "coef")
tmp %>% ggplot(aes(age, coef, color = region) ) + geom_smooth() + ggtitle("Impact of Median Age on region") + xlab("Median Age") + ylab("Probability of outcome")
```


```{r}

tmp = as.data.frame(cbind(mod3.0$fitted.values[,1], dat$medIncome))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,2], tmp))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,3], tmp))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,4], tmp))

names = c("Northeast",        "South"        ,"West" , "Midwest","age")
tmp = setNames(tmp, names)
tmp = tmp %>% pivot_longer(cols = !age, names_to = c("region" ), values_to = "coef")
tmp %>% ggplot(aes(age, coef, color = region) ) + geom_smooth() + ggtitle("Impact of Income on region") + xlab("Median Income") + ylab("Probability of outcome")
```

```{r}

tmp = as.data.frame(cbind(mod3.0$fitted.values[,1], dat$popEst2015))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,2], tmp))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,3], tmp))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,4], tmp))

names = c("Northeast",        "South"        ,"West" , "Midwest","age")
tmp = setNames(tmp, names)
tmp = tmp %>% pivot_longer(cols = !age, names_to = c("region" ), values_to = "coef")
tmp %>% ggplot(aes(age, coef, color = region) ) + geom_smooth() + ggtitle("Impact of population on region") + xlab("population") + ylab("Probability of outcome")

```

```{r}

tmp = as.data.frame(cbind(mod3.0$fitted.values[,1], dat$PctWhite))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,2], tmp))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,3], tmp))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,4], tmp))

names = c("Northeast",        "South"        ,"West" , "Midwest","age")
tmp = setNames(tmp, names)
tmp = tmp %>% pivot_longer(cols = !age, names_to = c("region" ), values_to = "coef")
tmp %>% ggplot(aes(age, coef, color = region) ) + geom_smooth() + ggtitle("Impact of percent of white population on region") + xlab("Percent of white population ") + ylab("Probability of outcome")

```

```{r}
tmp = as.data.frame(cbind(mod3.0$fitted.values[,1], dat$PctPrivateCoverage))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,2], tmp))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,3], tmp))
tmp = as.data.frame(cbind(mod3.0$fitted.values[,4], tmp))

names = c("Northeast",        "South"        ,"West" , "Midwest","age")
tmp = setNames(tmp, names)
tmp = tmp %>% pivot_longer(cols = !age, names_to = c("region" ), values_to = "coef")
tmp %>% ggplot(aes(age, coef, color = region) ) + geom_smooth() + ggtitle("Impact of percent of pubic insurance coverage on region") + xlab("Percent of") + ylab("Probability of outcome")
```


# Poisson Regression

### Over-dispersion
```{r}
hist(dat$TARGET_deathRate, freq = F, ylim = c(0, 0.04), main = "Density Distribution of Lung-Cancer Death Rate", xlab = "County Death Rate", ylab = "Density")
lines(as.integer(min(dat$TARGET_deathRate)):as.integer(max(dat$TARGET_deathRate)), dpois(as.integer(min(dat$TARGET_deathRate)):as.integer(max(dat$TARGET_deathRate)), lambda = mean(dat$TARGET_deathRate)))
```
```{r}
print(mean(dat$TARGET_deathRate))
print(var(dat$TARGET_deathRate))
```

```{r}
hist(dat$avgAnnCount/dat$popEst2015, freq = F, main = "Density Distribution of Average Annual Lung-Cancer Incidence Rate Per County", xlab = "Average Incidence Rate per County", ylab = "Density")#, ylim = c(0, 0.04))
lines(as.integer(min(dat$avgAnnCount/dat$popEst2015)):as.integer(max(dat$avgAnnCount/dat$popEst2015)), dpois(as.integer(min(dat$avgAnnCount/dat$popEst2015)):as.integer(max(dat$avgAnnCount/dat$popEst2015)), lambda = mean(dat$avgAnnCount/dat$popEst2015)))
```
```{r}
mean(dat$avgDeathsPerYear/dat$popEst2015)
var(dat$avgDeathsPerYear/dat$popEst2015)
```

```{r}
hist(dat$incidenceRate, freq = F, ylim = c(0, 0.02), main = "Density Distribution of Incidence Rate of All Cancers", xlab = "Incidence Rate of All Cancers (age-adjusted)", ylab = "Density")
lines(as.integer(min(dat$incidenceRate)):as.integer(max(dat$incidenceRate)), dpois(as.integer(min(dat$incidenceRate)):as.integer(max(dat$incidenceRate)), lambda = mean(dat$incidenceRate)))
```
```{r}
print(mean(dat$incidenceRate))
print(var(dat$incidenceRate))
```

### Model fits
```{r, warning = FALSE}
# poisson fit general (general = only include median income, state, population size)
pois_general <- dat %>% glm(formula = TARGET_deathRate ~ medIncome + State + popEst2015, family=poisson(), data=.)
summary(pois_general)

# neg bin fit general 
neg_bin <- dat %>% MASS::glm.nb(formula = TARGET_deathRate ~ medIncome + State + popEst2015, data=.)
summary(neg_bin)

neg_bin_EMM <- dat %>% MASS::glm.nb(formula = TARGET_deathRate ~ medIncome + State + popEst2015 + medIncome*popEst2015, data=.)
summary(neg_bin_EMM) # check effect modification 
# EMM modification since interaction term p-value = 0.043268 
# decrease in AIC keep EMM model (27078)
```


From linear regression model analysis, covariates included MedianAge, PctWhite, PctOtherRace, no_highschool, highschool, PctPublicCoverageAlone, PctPrivateCoverageAlone, PctPublicCoverageAlonexPctPrivateCoverageAlone, PercentMarried, PercentMarriedxMedianAge. 
Some of these covariates are considered for the poisson extension model:

```{r setup, warning=FALSE}
# NOTE: Some summary functions commented out to save space
# AIC and p-values for these were written as comments

# include median age
neg_bin_2 <- dat %>% MASS::glm.nb(formula = TARGET_deathRate ~ medIncome + State + popEst2015 + medIncome*popEst2015 + MedianAge, data=.)
summary(neg_bin_2)$aic # AIC decrease (27074)

# include race data 
neg_bin_3 <- dat %>% MASS::glm.nb(formula = TARGET_deathRate ~ medIncome + State + popEst2015 + medIncome*popEst2015 + MedianAge, data=.)
summary(neg_bin_3)$aic # AIC decrease (27074)

neg_bin_4 <- dat %>% MASS::glm.nb(formula = TARGET_deathRate ~ medIncome + State + popEst2015 + medIncome*popEst2015 + MedianAge + PctWhite + PctAsian + PctOtherRace, data=.)
summary(neg_bin_4)$aic # AIC decrease (27003)

neg_bin_3_EMM <- dat %>% MASS::glm.nb(formula = TARGET_deathRate ~ medIncome + State + popEst2015 + medIncome*popEst2015 + MedianAge + PctWhite + PctAsian + PctOtherRace + PctOtherRace*medIncome, data=.)
#summary(neg_bin_3_EMM) # check effect modification - only PctOtherRace*medIncome shown
# MedianAge*medIncome covar of interest not sig and not sig change in AIC
# PctWhite*medIncome covar of interest not sig and not sig change in AIC
# PctAsian*medIncome term is not sig and no sig change in AIC (slight increase)
# PctOtherRace*medIncome term is not sig and no sig change in AIC (slight increase)

# include education 
neg_bin_5 <- dat %>% MASS::glm.nb(formula = TARGET_deathRate ~ medIncome + State + popEst2015 + medIncome*popEst2015 + MedianAge + PctWhite + PctAsian + PctOtherRace + no_highschool + highschool, data=.)
summary(neg_bin_5)$aic # AIC decrease (26900), 
#but no_highschool (p= 0.256580), PctAsian (p=0.277823), popEst2015 (p=0.380916), 
      #and medIncome:popEst2015(p=0.362506) not significant

neg_bin_5b <- dat %>% MASS::glm.nb(formula = TARGET_deathRate ~ medIncome + State + MedianAge + PctWhite + PctAsian + PctOtherRace + highschool, data=.)
#summary(neg_bin_5b)
# remove no_highschool and popEst2015 (since sign for most of previous models 
    #have not been highly sig (p-value close to 0.05))
# AIC slightly decrease (26896) but almost no change, keep this model for simplicity
# PctAsian still not sig (p=0.418288)

neg_bin_5c <- dat %>% MASS::glm.nb(formula = TARGET_deathRate ~ medIncome + State + MedianAge + PctWhite + PctOtherRace + highschool, data=.)
summary(neg_bin_5c)$aic 
# AIC almost no change but ever so slight decrease (26895), 
#keep this model for simplicity

neg_bin_5c_EMM <- dat %>% MASS::glm.nb(formula = TARGET_deathRate ~ medIncome + State + MedianAge + PctWhite + PctOtherRace + highschool + highschool*medIncome, data=.)
#summary(neg_bin_5c_EMM) 
# check effect modification: no EMM (p=0.551039), keep neg_bin_5c

# include health coverage
neg_bin_6 <- dat %>% MASS::glm.nb(formula = TARGET_deathRate ~ medIncome + State + MedianAge + PctWhite + PctOtherRace + highschool + PctPublicCoverageAlone + PctPrivateCoverageAlone, data=.)
summary(neg_bin_6)$aic # AIC decrease a lot (21553)

neg_bin_6_EMM <- dat %>% MASS::glm.nb(formula = TARGET_deathRate ~ medIncome + State + MedianAge + PctWhite + PctOtherRace + highschool + PctPublicCoverageAlone + PctPrivateCoverageAlone + medIncome*PctPrivateCoverageAlone, data=.)
#summary(neg_bin_6_EMM) 
# check Effect modification; only medIncome*PctPrivateCoverageAlone term shown here
# PctPublicCoverageAlone*PctPrivateCoverageAlone: 
        ## AIC almost no change (21548); slight EMM, 
        ## but sig of other covariates seems to decrease a lot, so exclude
# PctPublicCoverageAlone*medIncome: no EMM (p=0.663161)
# medIncome*PctPrivateCoverageAlone: no EMM (p=0.895889)

# final model: neg_bin_6
neg_bin_final <- dat %>% MASS::glm.nb(formula = TARGET_deathRate ~ medIncome + State + MedianAge + PctWhite + PctOtherRace + highschool + PctPublicCoverageAlone + PctPrivateCoverageAlone, data=.)
summary(neg_bin_final)

# compare final model to regular poisson model with same covariates
pois_final <- dat %>% glm(formula = TARGET_deathRate ~medIncome + State + MedianAge + PctWhite + PctOtherRace + highschool + PctPublicCoverageAlone + PctPrivateCoverageAlone, family=poisson(), data=.)
summary(pois_final)
```

